Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2023-12-13 ms.qmd
More discussion writing

### **Acclimation capacity and opportunity for selection is limited in response to climate change**
One explanation for why terrestrial ectotherms show minimal acclimation capacity may be related to the fact that terrestrial ectotherms, were, on average acclimated for significantly less time than ectotherms from aquatic habitats. However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have higher rates of acclimation than fishes; See @Einum2023-lz]. Faster rates of acclimation would therefore result in the opposite patterns to those we observed -- in other words, terrestrial species would be more likely to exhibit lower $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ when controlling for acclimation time. 
assuming a fraction of this variation is heritable and/or it weakens selection

### **No signature of past climate on capacity for acclimation or opportunity for selection**

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key]. Higher spatial and temporal heterogenetiy in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely evolve, however, we did not find evidence to this effect; possibly because of the lower predictability of thermal environments over time making cues unreliable [@hendry2016key]. Empirical evidence properly testing these predictions is also scant. @Leung2020 recently showed that greater plasticty in morphological traits did evolve in experimental algae populations, however, work in seed beetles. Our results suggest that such theoretical explanations may not be general but rather specific to the population, context and trait in question; a point emphaised by others.

### ***Ecological consequences of reduced variation in physiological Rates***
Variance in physiological rates decreases substantially at high temperatures with effects being especially pronouced in terrestrial ecosystems and for key traits such as metabolism and performance. Changes in variability in physiological rates, particularly energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems that can have important consequences for population dynamics [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key]. For examaple, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population occilations compared to scenerios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ecotherms will likely face an onslaught of ecological and evolutionary challenges which will mean greater rates of extintion and reduced productivity – predictions already reinforced by a number of authorities . To date, however, an appreciation for the wide-reaching impacts changes in variance may have on populations and communities remains limited. Indeed, the ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remain unclear. Such insight will require the development of theoretical models that better integrate intrapopulation variation into mechanistic physiological models or ones that better capture stochasticity such as individual based models.

The dearth of empirical data demonstrating changes in variance paired with a lack of theoretical models 
- buckley 2014, - Lawton 1991, - Nisbet

### **Acute and acclimation responses of wild ectotherms are lower in warmer environments**

### **The opportunity for selection is expected to decrease for wild ectotherms under climate change**

### **Limitations and future directions**

An enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may increase our ability to make more informed predcitions about ecological and evolutonary dynamics of natural populations [@forsman2015rethinking].

>>>>

### ***Acclimation capacity and opportunity for selection across ectotherms in response to climate change***

There is a general interest in understanding the interplay between plasticity and genetic adaptation to a populations capacity to mitigate climate change effects []. We show that plastic adjustments to physiological rates are rather small (if they exist at all) compared to the general reduction variance in physiological rates. While it is unclear the extent to which such changes in variance are heritable we expect changes in variance of this magnitude to have an impact on the opportunity for selection by reducing the strength of selection or underlying genetic variance. Both are predcited to reduce the capacity of populations to evolve under climate change. One explanation for why terrestrial ectotherms show minimal acclimation capacity may be related to the fact that terrestrial ectotherms, were, on average acclimated for significantly less time than ectotherms from aquatic habitats. However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have higher rates of acclimation than fishes; See @Einum2023-lz]. Faster rates of acclimation would therefore result in the opposite patterns to those we observed -- in other words, terrestrial species would be more likely to exhibit lower $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ when controlling for acclimation time. 
assuming a fraction of this variation is heritable and/or it weakens selection

### ***No signature of past climate on capacity for acclimation***

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogenetiy in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely evolve in terrestrial environments, however, we did not find evidence to this effect.  The lack of relationship between acclimation responses and environmental variability and predictability could be due to the challenges in estimating realistic microthermal enviroments experienced by taxa or because lower predictability of thermal environments over time make cues unreliable preventing plastic responses from evolving easily [@Leung2020; @hendry2016key]. Randomly flutuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly flutuating thermal conditions did not lead to the expected increase in plasticity, but reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits when environments experienced were less predictable by experimentally evolving algae populations (*Dunaliella salina*) [see also @leung2023phenotypic]. The fact that aquatic ectotherms did show a capacity to acclimate suggest that environmental predcitability, which is likely generally higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

### ***Ecological consequences of reduced variation in physiological Rates***
While we found that variance in physiological rates decreases substantially at high temperatures, especially pronouced in terrestrial ecosystems, variance reductions in metabolism and performance were key traits most strongly impacted. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key]. Maintaining intrapopulaton variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. Nonetheless, few studies have explored the consequences that variation in physiological traits have for population dynamics. Those studies that have have shown that such variation can have important implications. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population occilations compared to scenerios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ecotherms will likely face an onslaught of ecological and evolutionary challenges which will mean greater rates of extintion and reduced productivity – predictions already reinforced by a number of studies [@hoffmann2013upper; @Gunderson2015; @Pottier2022]. To date, however, an appreciation for the wide-reaching impacts that changes in variance may have on populations and communities remains limited. The ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remain unclear. What we desperately need now are the development of theoretical models that better integrate intrapopulation variation into mechanistic physiological models or ones that better capture stochasticity such as individual based models. Our meta-analytic findings can help us better position ourselves to now start asking questions about the ecological and evolutionary consequences of changes in variance in physiological rates.

### ***Limitations and future directions***

An enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may increase our ability to make more informed predictions about ecological and evolutonary dynamics of natural populations [@forsman2015rethinking]. Variance in physiological rates could be better incoperated in physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. Our meta-analysis mean that such models can be that are better aligned with empirical findings. Nonethless, many fascinating questions remain unanswered that will require greater focus on the consequences of change sin variance (rather than just the mean): How does a reduction in physiological rate varaince change the flow of energy across tropic levels within communities? What are the biochemical, cellular and physiological mechanisms that underly reduced variance in physiological rates? What fraction of the change in variance is environmental versus genetic?
